Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021

### 2024-01-25 outputs/paper/paper.qmd
update tables, graphs and results

---
title: "Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021"
title-block-banner: true
abstract: "Police strip search has been a topic of great contention during the COVID-19 pandemic. The question commonly raised was how certain demographics are more likely to undergo arrests and strip searches than others, and why has this skewed power dynamics been maintained for so long. Data then was gathered from Open Data Toronto to assess the correlation between strip searches and the racial and gender identities of arrested 'suspects' in the Toronto Municipal from 2020 to 2021. This analysis would enable us to rethink the modern form of policing, how it transpires in times of crisis and critically affects people of marginalized identities. More on RESULTS" 
thanks: "Code and data are available at: https://github.com/ponolite/Police_Arrests_and_Strip_Searches/tree/main"
author: "Quang Mai"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(opendatatoronto)
library(dplyr)
library(readr)
library(kableExtra)
library(here)
library(ggbeeswarm)
library(ggrepel)
library(ggplot2)

@@

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism, contributing to the dialogue around police power abuse, using strip-search to undermine the the vulnerable populations in Toronto. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using stacked summary tables, stacked bar charts and bubble charts. Particularly, for each category of race and gender, the distribution bars or points would be sectioned into various categories like 'including strip searches' or 'not including' to visualize the statistics of strip searching for intersecting social identities With that, It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Yet, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

This chosen dataset was published by Toronto Police Services to convey information related to all police arrests and strip searches from 2020 to 2021 (last updated on December 2, 2022). Originally, the dataset features information on date of arrest, event id, the arrested person's race, sex, and age group, arrest location, whether the arrest includes strip searching, the person's action at the time of arrest, and the police's reason for strip searching along with any items found. Data of strip searches are lodged based on each police arrest throughout 2020 and 2021, using a unique identifier called `event_id` which identifies the nature of the strip search and the social identities of the people involved.

Informative as it is, the dataset contains much naming inconsistency and many columns outside the scope of this paper. Thus, in order to conduct an intersectional---or gender and race---analysis of police strip searches, I have to further clean this dataset and split it into two smaller datasets. Additional details on data extraction and cleaning will be specified in subsequent sections.

@@

The other dataset conveys all police arrests that involve strip searching and the reasons behind strip searching. Under close examination, the naming of individual data cells in different columns from the original, raw dataset is quite disordered, for instance, using `None` and `XX` in place of the usual `NA`. Thus, all inconsistent data cells were renamed to streamline the naming conventions and data processing later on. After renaming, the big dataset was split into containing relevant columns ([see @tbl-cleaned_search_reasons_data]). Then, for this particular dataset, all instances of police arrests without strip searching are dropped using the `filter()` function, along with a `!is.na()` function for rows with `NA` as values for all the `reason` column.

Except for "ID" column, all other columns in this new dataset are reasons which account for why the Toronto police force conducts strip search under certain arrests. For instance, "Reason - Assist Escape" is for people who may have helped other suspects escape and "Reason - Possess Evidence" is for people who are suspected of harboring items like drugs or an alibi. For each "Reason" column, if the data cell contains `0`, this means that strip search doesn't happen because of that particular reason. Vice versa, if it contains `1`, this means strip search happens because of that reason. In addition, a new column called "Total Reasons" totals all reasons used to account for each police strip search is added with `mutate()`, where all instances of `1` and `0` are changed into numeric with `as.numeric()` and then summed together.

@@

Furthermore, the reason for keeping rows of events that don't include strip search is to later produce summary statistics of events with and without strip searches. This further aligns with the task of the paper at hand, which is to critically reflect on the rate of strip searching on people with racialized and gendered identities.

@@

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  group_by(race) |>
  mutate(percentage = round(number_of_arrests/sum(number_of_arrests)*100)) |>
  kable(
    col.names = c("Race", "Strip Search", "Number of Arrests", "Percentage"),
    booktabs = TRUE)

@@

```{r fig.height=6, fig.width=7}
#| echo: false
#| message: false
#| label: fig-race_strip_search
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

  cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  ggplot(aes(x = race, y = number_of_arrests, fill = strip_search)) + 
  geom_bar(stat="identity") +
  theme(legend.key.size = unit(0.5, 'cm'), axis.text.x = element_text(angle = 70, vjust = 1, hjust= 1)) +
  labs(x = "Race",
       y = "Number of Arrests",
       fill = "Strip Search") + scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  filter(strip_search == "Yes") |> 
  ggplot(mapping = aes(x = strip_search, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar() + 
  labs(y = "Number of Arrests", x = "Strip Search", fill = "Strip Search") +
  facet_wrap(~race) + scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  group_by(gender) |>
  mutate(percentage = round(number_of_arrests/sum(number_of_arrests)*100)) |>
  kable(
    col.names = c("Gender", "Strip Search", "Number of Arrests", "Percentage"),
    booktabs = TRUE)

@@

  cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(number_of_arrests = n) |>
  ggplot(aes(x = gender, y = number_of_arrests, fill = strip_search)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Gender",
       y = "Number of Arrests",
       fill = "Strip Search"
  ) + scale_fill_viridis_d(option="cividis")

@@

```{r}
#| echo: false
#| message: false
#| results: hide
#| label: tbl-reasons-strip-search
#| tbl-cap: Summary Statistics of Items Found During Police Strip Searches

cleaned_race_gender_data |>
  group_by(strip_search, items_found) |>
  count() |>
  rename(number_of_arrests = n) |>
  filter(strip_search == "Yes") |>
  select(items_found, number_of_arrests) |>
  kable(
    col.names = c("Strip Search", "Items Found", "Number of Arrests"),
    booktabs = TRUE
  )
```

```{r}
#| echo: false
#| message: false
#| label: tbl-items-strip-search
#| tbl-cap: Summary Statistics of Reasons for Police Strip Searches

cleaned_search_reasons_data |>
  group_by(total_reasons, items_found) |>
  count() |>
  rename(count = n) |>
  kable(
    col.names = c("Number of Reasons", "Items Found", "Count"),
    booktabs = TRUE
  )
```

@@

>>>>

---
title: "Slight Positive Correlation Between Gendered and Racialized Police Strip Searches in the Toronto Municipal, 2020-2021"
title-block-banner: true
abstract: "Police strip search has been a topic of great contention during the COVID-19 pandemic. The question commonly raised was how certain demographics are more likely to undergo arrests and strip searches than others, and why has this skewed power dynamics been maintained for so long. Data then was gathered from Open Data Toronto to assess the correlation between strip searches and the racial and gender identities of arrested 'suspects' in the Toronto Municipal from 2020 to 2021. This analysis would enable us to rethink the modern form of policing, how it transpires in times of crisis and critically affects people of marginalized identities. More on RESULTS" 
thanks: "Code and data are available at: https://github.com/ponolite/Police_Arrests_and_Strip_Searches/tree/main"
author: "Quang Mai"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
output: 
  pdf_document :
    keep_tex: true
    number_sections: true
header-includes:
 \usepackage{wrapfig}
 \usepackage{float}
 \floatplacement{figure}{H}
---

@@

library(tidyverse)
library(knitr)
library(opendatatoronto)
library(dplyr)
library(readr)
library(kableExtra)
library(here)
library(ggrepel)
library(ggplot2)
library(float)

citation("float")

@@

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism, contributing to the dialogue around police power abuse, using strip-search to undermine the the vulnerable populations in Toronto. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using stacked summary tables, stacked histograms and bubble charts. Particularly, for each category of race and gender, the distribution bars or points would be sectioned into various categories like 'including strip searches' or 'not including' to visualize the statistics of strip searching for intersecting social identities With that, It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Yet, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

Published by Toronto Police Services, this dataset conveys information related to all police arrests and strip searches from 2020 to 2021 (last updated on December 2, 2022). Originally, the dataset features 26 varibles, including date of arrest, event id, the arrested person's race, sex, and age group, arrest location, whether the arrest includes strip searching, the person's action at the time of arrest, and the police's reason for strip searching along with any items found. Data of strip searches are lodged based on each police arrest throughout 2020 and 2021, using a unique identifier called `event_id` which identifies the nature of the strip search and the social identities of the people involved.

Informative as it is, the dataset contains much naming inconsistency and many variables outside the scope of this paper. Thus, in order to conduct an intersectional---or gender and race---analysis of police strip searches, I have to further clean this dataset and split it into two smaller datasets. Additional details on data extraction and cleaning will be specified in subsequent sections.

@@

To further inspect the nature of each strip search, reasons for strip searching prove a vital data source. Thus, this dataset--another sub-dataset split from the orginal raw data--conveys all police arrests that involve strip searching and the reasons behind strip searching. Under close examination, the naming of individual data cells from the original, raw dataset is quite disordered, for instance, using `None` and `XX` in place of the usual `NA`. Thus, all inconsistent data cells were renamed to streamline the naming conventions and data processing later on. After renaming, the big dataset was refined into containing relevant columns ([see @tbl-cleaned_search_reasons_data]). Then, for this particular dataset, all instances of police arrests without strip searching are dropped using the `filter()` function, along with a `!is.na()` function for rows with `NA` as values for all of the `reason` variables

Except for "ID" variable, all other variables in this new dataset are reasons to account for why the Toronto police force conducts strip search under certain arrests. For instance, "Reason - Assist Escape" is for people who may have helped other suspects escape and "Reason - Possess Evidence" is for people who are suspected of harboring items like drugs or an alibi. For each "Reason" variable, if the data cell contains `0`, this means that strip search doesn't happen. Vice versa, if it contains `1`, this means strip search happens. In addition, a new column called "Total Reasons" is added to sum  all reasons used to account for each strip search. To do this, `mutate()` is utilized, where all instances of `1` and `0` are changed into numeric with `as.numeric()` and then summed together.

@@

To encapsulate, throughout 2020 and 2021, there are a total of 7801 recorded strip searches out of 65276 incidents of police arrests, which accounts for a whooping 12% of all arrests (see [@tbl-items-strip-search]). Excluding those whose racial identities are unknown and non-existent, by percentage, the three most affected racial groups are, in descending order, of Indigenous, Black and White ethnic roots, standing at 16%, 14% and 13% each (see [@tbl-race-strip-search]). Gender-wise, men are more likely to be strip searched compared to women, at 12% and 10% respectively (see [@tbl-gender-strip-search]). Though, observably, the difference is quite marginal. To note, people of unidentified gender identity aren't strip searched at all according to the dataset (denoted as `U` in the tables and graphs).

It's worth noting that to account for the discrepancies between counts of strip search grouped by different racial identities are challenging. This is due to the incongruous population spread of each racial group within Toronto. For instance, Toronto is largely made up of people from Caucasian descent, thus it's more probable that the people arrested and strip searched by the police force are largely white in the dataset (see [@@tbl-race-strip-search]) Therefore, to render the data statistically proportionate to one another,

@@

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(race) |>
  mutate(percentage = round(count/sum(count)*100),
         total_arrests = sum(count)) |>
  kable(
    col.names = c("Race", "Strip Search", "Count", "Percentage", "Total Arrests"),
    booktabs = TRUE)

@@

In addition, a new column called "Total Reasons" totals all reasons used to account for each police strip search is added with `mutate()`, where all instances of `1` and `0` are changed into numeric with `as.numeric()` and then summed together.
to ensure that all data are comparative to one another

```{r}
#| echo: false
#| message: false
#| label: fig-race_strip_search
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(race) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(aes(x = race, y = percentage, fill = strip_search)) + 
  geom_bar(stat="identity") +
  coord_flip() +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Race",
       y = "Percentage",
       fill = "Strip Search") + scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(race, strip_search) |>
  count() |>
  rename(count= n) |>
  group_by(race) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(mapping = aes(x = gender, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar() + 
  labs(y = "Percentage", x = "Strip Search", fill = "Strip Search") +
  facet_wrap(~race) + scale_fill_viridis_d(option="cividis")

@@

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(gender) |>
  mutate(percentage = round(count/sum(count)*100),
         total_arrests = sum(count)) |>
  kable(
    col.names = c("Gender", "Strip Search", "Count", "Percentage", "Total Arrests"),
    booktabs = TRUE)

@@

  cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = gender, y = count, fill = strip_search)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Gender",
       y = "Count",
       fill = "Strip Search"
  ) + scale_fill_viridis_d(option="cividis")

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(gender) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(aes(x = gender, y = percentage, fill = strip_search)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Gender",
       y = "Percentage",
       fill = "Strip Search"
  ) + scale_fill_viridis_d(option="cividis")

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(gender) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(mapping = aes(x = strip_search, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar() + 
  labs(y = "Count", x = "Strip Search", fill = "Strip Search") +
  facet_wrap(~gender) + scale_fill_viridis_d(option="cividis")

@@

As for reasons of strip searching provided by the Toronto Police Services and items found along these strip searches, there is a substantial 10% of strip searches being conducted without any reason (see [@tbl-reasons-strip-search]). Most commonly, strip searches are carried out under at least 1 or 2 reasons, both at 29% (see [@tbl-reasons-strip-search]). Overall, most of strip search enforcement doesn't garner any items in question from the people who were suspected by the police. Specifically, a whooping 63% of strip searches don't have any items found (see [@tbl-items-strip-search]).

```{r}
#| echo: false
#| message: false
#| label: tbl-items-strip-search
#| tbl-cap: Summary Statistics of Items Found During Police Strip Searches

cleaned_race_gender_data |>
  group_by(strip_search, items_found) |>
  count() |>
  rename(count = n) |>
  filter(strip_search == "Yes") |>
  select(items_found, count) |>
  group_by(strip_search) |>
  mutate(total_searches = sum(count),
         percentage_of_items_found = round(count/total_searches*100)) |>
  kable(col.names = c("Strip Search", "Items Found", "Count", "Total Searches", "Percentage of Items Found"), booktabs = TRUE
        )
```

```{r}
#| echo: false
#| message: false
#| label: tbl-reasons-strip-search
#| tbl-cap: Summary Statistics of Reasons for Police Strip Searches

cleaned_search_reasons_data |>
  group_by(total_reasons, items_found) |>
  count() |>
  rename(count = n) |>
  group_by(total_reasons) |>
  mutate(percentage = round(count/sum(count)*100),
         total_searches = sum(count),
         percentage_total = round(total_searches/7801*100)) |>
  kable(
    col.names = c("Total Reasons", "Items Found", "Count", "Percentage by Reasons", "Total Searches", "Percentage by Searches"),
    booktabs = TRUE
  ) |> 
  kable_styling(latex_options="scale_down")
```

@@

binaristic view of gender

mystified, how is race sectioned and divided

biased source, observation
